A Domain-Specific Optimizing Compiler\\for Solving Partial Differential Equations

### 2014-02-22 paper/bare_conf.tex
Huge refactor of intro, background, compiler

\section{Introduction}
In many fields, like computational fluid dynamics, computational electromagnetics, and structural mechanics, programs model phenomena by means of partial differential equations (PDEs). Numerical techniques, like Finite Volume Method (FVM) and Finite Element Method (FEM), are widely employed in real-world applications to approximate solutions of PDEs. Unstructured meshes are often used to discretize the domain of a PDE, since their geometric flexibility allows solvers to be extrimely effective. However, they are also one of the main responsible for performance degradation, due to the need for indirect accesses (e.g. \texttt{A[B[i]]}) to access mesh components. A typical problem, therefore, is that implementing a scientific application requires a great effort, other than domain knowledge, and even more time is needed to optimize their execution time.

An unstructured mesh discretizes the domain of the equation into two- or three-dimensional cells (or ``elements''), within which the solution is sought. As the number of elements can be of the order of millions, a major issue is the time required to perform the numerical method, which can be hours or days. To address this issue, domain-specific languages (DSLs), which ease the programming burden and enable highly-efficient parallel execution, have been developed. The successful porting of Hydra, a CFD industrial application devised by Rolls Royce for turbomachinery design (based on FVM, roughly 50000 lines of code and mesh sizes that can be over 100 millions edges), to the OP2 DSL~\cite{pyop2ics}, demonstrates the effectiveness of the DSL approach to implementing PDEs solvers~\cite{OP2-hydra}.

OP2, as well as Lizst~\cite{lizst}, adopt a kernel-oriented programming model, in which the computation semantics is expressed through self-contained functions (or ``kernels''). A kernel is applied to all elements in a set of mesh components (e.g. edges, vertices, elements), with an implicit synchronization between the application of two consecutive kernels. On commodity multi-cores, a kernel is executed sequentially by a thread, while parallelism is achieved partitioning the mesh and assigning each partition to a thread. Together with the indirections problem, kernel optimization is one of the major concerns in unstructured mesh applications. In this paper, we tackle this problem by proposing a domain-driven optimization strategy for a class of kernels used in FEM.

We focus on Finite Element Local Assembly (``assembly'', in the following), a fundamental step of FEM that covers an important fraction of the overall computation run-time, typically in the range 30$\%$-60$\%$. During the assembly phase, the solution of the PDE is approximated by executing a suitable kernel over all elements in the discretized domain. A kernel's working set is usually small enough to fit the L1 cache; it might need L2 cache when high-order methods are employed to improve the accuracy of the solution. However, we do not consider the latter case. An assembly kernel is characterized by the presence of an affine, usually non-perfect loop nest, where individual loops are rather small (the trip count rarely exceeds 60, with a minimum value of 3, depending on the order of the method). With such small kernels, our work focuses on aspects like minimization of floating-point operations, register allocation and instruction-level parallelism, especially in the form of SIMD vectorisation.

Our study is conducted in the context of Firedrake, a multilayered system for the automated solution of PDEs through FEM. PDEs are written in Firedrake using a domain-specific language called Unified Form Language (UFL). The high-level UFL notation is used by the Fenics Form Compiler (FFC) to produce assembly kernels. At the core of Firedrake~\cite{firedrake} there is PyOP2~\cite{pyop2ics} - a python implementation of the OP2 abstraction - which triggers compilation of kernels using an available vendor compiler, and eventually manages parallel execution over the unstructured mesh. We exploit the domain knowledge and the structure inherent to FFC's assembly kernels to develop an optimizing compiler for all problems that are expressible in Firedrake.

The compiler we have developed is integrated with PyOP2, and already used in the released version of Firedrake~\cite{firedrake-code}. This allows us to evaluate the proposed code transformations in a range of real-world problems, employing different low-order methods (from $p=1$ to $p=4$; the higher, the more expensive is a kernel, but better is the accuracy), and varying the topology of the mesh (2D, 3D, 3D hybrid structured-unstructured, which in turn affect the kernel cost). Performance results show that compiler-generated code for non-trivial assembly kernels, if not adequately transformed, is sub-optimal. A cost-model-driven sequence of source-to-source code transformations, aimed at improving register allocation and register data locality, can result in performance improvements up to 33$\%$ over ``softly-optimized'' code (i.e. where only basic transformations are performed, such as auto-vectorizable loop-invariant code motion, padding and data alignment), and up to 77$\%$ over the original FFC kernels. The contribution of this paper is therefore twofold:
\begin{itemize}
\item An optimisation strategy for a class of kernels widely-used in scientific applications, namely Finite Element Local Assembly kernels. Our approach exploits domain knowledge to go beyond the limits of both vendor (e.g. $icc$) and research (e.g. those based on the polyhedral model) compilers.
\item Automation of such code optimizations in Firedrake, a framework for the resoloution of PDEs through FEMs. This enables performing an in-depth performance evaluation of the proposed optimization strategy in real-world simulations.
\end{itemize}

@@

\section{Background}
%Here we discuss about the computational characteristics of computational science kernels. Briefly cite op2 kernels. Emphasis on Finite Element Assembly. Generalization and formalization of a Finite Element Assembly kernel using quadrature representation.
\begin{algorithm}[t]
\small
\caption{Local assembly code generated by FFC for a Helmholtz problem (2D mesh, Lagrange $p=1$ elements).}
\label{code:helmholtz}
\KwSty{void} helmholtz(double A[3][3], double **coords) $\lbrace$\\
~~// K, det = Compute Jacobian (coords) \\
~~\\
~~\KwSty{static const double} W3[3] = $\lbrace$...$\rbrace$\\
~~\KwSty{static const double} X$\_$D10[3][3] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\KwSty{static const double} X$\_$D01[3][3] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$3; i++) \\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$3; k++) \\
~~~~~~~~A[j][k] += ((Y[i][k]*Y[i][j]+\\
~~~~~~~~~~~+((K[1]*X$\_$D10[i][k]+K[3]*X$\_$D01[i][k])*\\
~~~~~~~~~~~*(K[1]*X$\_$D10[i][j]+K[3]*X$\_$D01[i][j]))+\\
~~~~~~~~~~~+((K[0]*X$\_$D10[i][k]+K[2]*X$\_$D01[i][k])*\\
~~~~~~~~~~~*(K[0]*X$\_$D10[i][j]+K[2]*X$\_$D01[i][j])))*\\
~~~~~~~~~~~*det*W3[i]);\\
$\rbrace$
\end{algorithm}

@@

Local assembly consists of evaluating so called element stiffness matrix and vector; in this work, we focus on computation of matrices, which is the costly part of the process. A stiffness matrix can be intuitively thought as an approximated representation of the PDE solution in a specific cell of the discretized domain. Quadrature algorithms (based on numerical integration), which follow on from the mathematical description of a stiffness matrix, are widely-used in scientific codes to implement assembly~\cite{quadrature1, quadrature2}. The FFC compiler used by Firedrake and Fenics can generate assembly kernels using optimized quadrature schemes. It is worth noticing, however, that problems and considerations similar to that exposed below can be found also in non-FFC assembly codes.

The complexity of FFC-generated kernels depends on the mathematical problem being solved. In simpler cases, the loop nest is perfect, it has short trip counts (in the range 3-15), and the computation reduces to a summation of a few, small outer-product vector-operations. An example is provided in Figure~\ref{code:helmholtz}, which shows an assembly kernel for a Helmholtz problem, using Lagrange basis functions on 2D elements with polynomial order $p=1$. In other scenarios, for instance when solving a non-linear problem like Burgers (see Figure~\ref{code:burgers}), the number of arrays involved in the computation of $A$ can be much larger: in this case, 14 unique arrays are accessed, and the same array can be referenced multiple times within the expression. Also, constants evaluated in outer loops (called $F$ in the code), acting as scaling factors of arrays, may be required; trip counts can be larger (proportionally to the order of the method); arrays may be block-sparse. Despite such a big space of possible assembly implementations, it is still possible to identify a domain-specific structure common to all kernels, which can be turned into effective code transformations and simd-vectorization.

@@

One attractive possibility is to transform the loop nest into a sequence of calls to BLAS routines. There are several issues that make achieving this difficult and, potentially, not effective. The first problem concerns automating identification and extraction of matrix-matrix multiplications from the code, which requires a non-trivial analysis of the memory access pattern. Related to this problem is the fact that, in general, many sub-expressions should be pre-computed (i.e. computed once before the loop nest), which leads to using temporaries and, possibly, to working sets that do not fit the L1 cache (for $p>2$). In addition, using BLAS routines may impact data locality, since the same matrix can appear in multiple products, implying, at least, a certain degree of register spilling. Finally, it is known that BLAS routines perform far from peak performance when the dimension of the matrices is rather small~\cite{nek5000}. Hand-made implementations of the simple Helmholtz problem ($p=1,2,3,4$) as a sequence of BLAS ($dgemm$) routines have demonstrated that performance is worse than that shown in Section~\ref{sec:perf-results}. A comprehensive study concerning BLAS optimization is therefore left as further work.

@@

\begin{algorithm}[t]
\small
\caption{Local assembly code generated by FFC for a Helmholtz problem (2D mesh, Lagrange $p=1$ elements). The padding, data alignment and $licm$ optimizations are applied. Data alignment and padding relate to an AVX machine. In this specific case, sub-expressions invariant to $j$ are identical to those invariant to $k$; in general, this could not be the case.}
\label{code:helmholtz-licm}
\KwSty{void} helmholtz(double A[3][4], double **coords) $\lbrace$\\
~~// K, det = Compute Jacobian (coords) \\
~~\\
~~\KwSty{static const double} W3[3] = $\lbrace$...$\rbrace$\\
~~\KwSty{static const double} X$\_$D10[3][4] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\KwSty{static const double} X$\_$D01[3][4] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$3; i++) $\lbrace$ \\
~~~~double LI$\_$0[4];\\
~~~~double LI$\_$1[4];\\
~~~~\KwSty{for} (\KwSty{int} r = 0; r$<$4; r++) $\lbrace$ \\
~~~~~~LI$\_$0[r] = ((K[1]*X$\_$D10[ip][r])+(K[3]*X$\_$D01[ip][r]));\\
~~~~~~LI$\_$1[r] = ((K[0]*X$\_$D10[ip][r])+(K[2]*X$\_$D01[ip][r]));\\
~~~~$\rbrace$\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
~~~~~~~~A[j][k] += (Y[i][k]*Y[i][j]+LI$\_$0[k]*LI$\_$0[j]+\\
~~~~~~~~~~~~~~~~~~~~+LI$\_$1[k]*LI$\_$1[j])*det*W3[i]);\\
~~$\rbrace$\\
$\rbrace$
\end{algorithm}

\begin{algorithm}[t]
\small
\caption{Local assembly code generated by FFC for a Helmholtz problem (2D mesh, Lagrange $p=1$ elements). The padding, data alignment, $licm$ and $split$ optimizations are applied. Data alignment and padding relate to an AVX machine. In this specific case, the split factor is 2.}
\label{code:helmholtz-split}
\KwSty{void} helmholtz(double A[3][4], double **coords) $\lbrace$\\
~~// K, det = Compute Jacobian (coords) \\
~~// Declaration of basis function matrices \\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$3; i++) $\lbrace$ \\
~~~~double LI$\_$0[4];\\
~~~~double LI$\_$1[4];\\
~~~~\KwSty{for} (\KwSty{int} r = 0; r$<$4; r++) $\lbrace$ \\
~~~~~~LI$\_$0[r] = ((K[1]*X$\_$D10[ip][r])+(K[3]*X$\_$D01[ip][r]));\\
~~~~~~LI$\_$1[r] = ((K[0]*X$\_$D10[ip][r])+(K[2]*X$\_$D01[ip][r]));\\
~~~~$\rbrace$\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
~~~~~~~~A[j][k] += (Y[i][k]*Y[i][j]+LI$\_$0[k]*LI$\_$0[j])*det*W3[i];\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
~~~~~~~~A[j][k] += LI$\_$1[k]*LI$\_$1[j]*det*W3[i];\\
~~$\rbrace$\\
$\rbrace$
\end{algorithm}

@@

\begin{algorithm}[t]
\small
\caption{Local assembly code generated by FFC for a Helmholtz problem (2D mesh, Lagrange $p=2$ elements). The padding, data alignment, $licm$ and $op$-$vect$ optimizations are applied. Data alignment and padding relate to an AVX machine. In this specific case, the unroll-and-jam factor is 1.}
\label{code:helmholtz-opvect}
\KwSty{void} helmholtz(double A[8][8], double **coords) $\lbrace$\\
~~// K, det = Compute Jacobian (coords) \\
~~// Declaration of basis function matrices \\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$6; i++) $\lbrace$ \\
~~~~// Do loop-invariant code motion \\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$4; j+=4) \\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$8; k+=4) $\lbrace$\\
~~~~~~~~// Call $Load$ and $set$ intrinsics \\
~~~~~~~~// Compute A[1,1],A[2,2],A[3,3],A[4,4] \\
~~~~~~~~// One $permute\_pd$ intrinsics per $k$-loop $load$\\
~~~~~~~~// Compute A[1,2],A[2,1],A[3,4],A[4,3] \\
~~~~~~~~// One $permute2f128\_pd$ intrinsics per $k$-loop $load$\\
~~~~~~~~// ...\\
~~~~~~$\rbrace$\\
~~~~// Do Remainder loop (from $j=4$ to $j=6$)\\
~~$\rbrace$\\
~~// Restore the storage layout:\\
~~\KwSty{for} (\KwSty{int} j = 0; j$<$4; j+=4) $\lbrace$\\
~~~~$\_\_$m256d r0, r1, r2, r3, r4, r5, r6, r7;\\
~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$8; k+=4) $\lbrace$\\
~~~~~~r0 = $\_$mm256$\_$load$\_$pd ($\&$A[j+0][k]);\\
~~~~~~// Load A[j+1][k], A[j+2][k], A[j+3][k]\\
~~~~~~r4 = $\_$mm256$\_$unpackhi$\_$pd (r1, r0);\\
~~~~~~r5 = $\_$mm256$\_$unpacklo$\_$pd (r0, r1);\\
~~~~~~r6 = $\_$mm256$\_$unpackhi$\_$pd (r2, r3);\\
~~~~~~r7 = $\_$mm256$\_$unpacklo$\_$pd (r3, r2);\\
~~~~~~r0 = $\_$mm256$\_$permute2f128$\_$pd (r5, r7, 32);\\
~~~~~~r1 = $\_$mm256$\_$permute2f128$\_$pd (r4, r6, 32);\\
~~~~~~r2 = $\_$mm256$\_$permute2f128$\_$pd (r7, r5, 49);\\
~~~~~~r3 = $\_$mm256$\_$permute2f128$\_$pd (r6, r4, 49);\\
~~~~~~$\_$mm256$\_$store$\_$pd ($\&$A[j+0][k], r0);\\
~~~~~~// Store A[j+1][k], A[j+2][k], A[j+3][k]\\
~~~~$\rbrace$\\
~~$\rbrace$\\
$\rbrace$
\end{algorithm}

@@

\begin{table}[h]
\begin{tabulary}{1.0\columnwidth}{|C|C|}
\hline
Name (Abbreviation) & Parameter \\\hline\hline
Auto-vectorizable Loop-invariant code motion ($licm$) &   \\ \hline
Padding &  \\ \hline
Data Alignment & \\ \hline
Loop interchange      & loops  \\ \hline
Loop unrolling  & unroll factor \\ \hline
Register tiling & tile size \\ \hline
Outer-product vectorization ($op$-$vect$) & tile size \\ \hline
Assembly splitting ($split$) & split point, split factor \\ \hline
\end{tabulary}
\caption{Overview of code transformations for FFC-generated assembly kernels.}
\label{table:code-transformations}
\end{table}

@@

Our approach to solving partial differential equations is based on domain-specific languages and libraries. Fenics and Firedrake provide the high-level Unified Form Language (UFL) to write problems in a notation resembling mathematical equations (``forms''). In Fenics, this specification is translated by a form compiler (FFC) into C code, which is subsequently just-in-time compiled and executed on the target architecture. In Firedrake we adopt a completely different approach. We think that FFC should produce a high-level representation of a Finite Element assembly kernel, and should abstract from the specific details of the underlying architecture. We have modified FFC, therefore, to build Abstract Syntax Trees (AST), rather than C code. ASTs are then passed to PyOP2, which implements the core of Firedrake, where parallel execution over the unstructured mesh is managed. We have developed a compiler that, given an AST conformed to the PyOP2 specification (such as those produced by the new FFC), can apply the transformations described in Section~\ref{sec:code-transf} and, subsequently, generate C code. The C representation is eventually just-in-time compiled for one of the backends that are currently supported by PyOP2, namely OpenMP, MPI, OpenMP+MPI, CUDA, OpenCL. Our compiler is now fully integrated with PyOP2, so it supports any kind of problem expressible in UFL.

>>>>

\section{Introduction}
In many fields, like computational fluid dynamics, computational electromagnetics, and structural mechanics, programs model phenomena by means of partial differential equations (PDEs). Numerical techniques, like Finite Volume Method and Finite Element Method, are widely employed in real-world applications to approximate solutions of PDEs. Unstructured meshes are often used to discretize the equation domain, since their geometric flexibility allows solvers to be extremely effective. The solution is sought in each cell of the discretized domain by applying suitable numerical kernels. Iteration over mesh components is also expensive, due to the presence of indirect memory accesses. As the number of cells can be of the order of millions, a major issue is the time required to execute the computation, which can be hours or days. To address this issue, domain-specific languages (DSLs) have been developed. The successful porting of Hydra, a computational fluid dynamics industrial application devised by Rolls Royce for turbomachinery design (based on Finite Volume Method, roughly 50000 lines of code and mesh sizes that can be over 100 millions edges), to OP2~\cite{pyop2ics}, demonstrates the effectiveness of the DSL approach for implementing PDEs solvers~\cite{OP2-hydra}.

OP2 adopts a kernel-oriented programming model, in which the computation semantics is expressed through self-contained functions (``kernels''). A kernel is applied to all elements in a set of mesh components (e.g. edges, vertices, elements), with an implicit synchronization between the application of two consecutive kernels. On commodity multi-cores, a kernel is executed sequentially by a thread, while parallelism is achieved partitioning the mesh and assigning each partition to a thread. Together with the indirections problem, kernel optimization is one of the major concerns in unstructured mesh applications. In this paper, we tackle this problem by proposing a domain-driven optimization strategy for a class of kernels used in Finite Element Methods.

We focus on Local Assembly (``assembly'', in the following), a fundamental step of a Finite Element Method that covers an important fraction of the overall computation run-time, typically in the range 30$\%$-60$\%$. During the assembly phase, the solution of the PDE is approximated by executing a suitable kernel over all elements in the discretized domain. A kernel's working set is usually small enough to fit the L1 cache; it might need L2 cache when high-order methods are employed to improve the accuracy of the solution. However, we do not consider the latter case. An assembly kernel is characterized by the presence of an affine, usually non-perfect loop nest, where individual loops are rather small (the trip count rarely exceeds 40, with a minimum value of 3, depending on the order of the method). With such small kernels, we focus on aspects like minimization of floating-point operations, register allocation and instruction-level parallelism, especially in the form of SIMD vectorisation.

Since loop nests are affine, one possibility is to rely on polyhedral compilation for improving performance. There are, however, several limitations. The first problem is that a polyhedral compiler reasons on iteration spaces, whereas some of the transformations we propose rely on analysis of the data space. Our attempts at using PoCC's optimizations~\cite{PoCC}, especially register tiling, were unsuccessful. A domain-specific transformation of the iteration space, which is not expressible in the polyhedral model because of its non-affine nature, also showed a significant performance gain due to improved register locality.

Assembly kernels could be turned into a sequence of calls to BLAS routines, although it is complicated to achieve and, potentially, ineffective. One issue concerns automating identification and extraction of matrix-matrix multiplications from the code, which requires a non-trivial analysis of the memory access pattern. In addition, register locality would be affected, since the same matrix can appear in a number of products. Finally, it is known that BLAS routines perform far from peak performance when the dimension of the matrices is small~\cite{nek5000}. Hand-made implementations of the simple Helmholtz assembly kernel as a sequence of BLAS $dgemm$ calls, for a range of polynomial orders ($p=1,2,3,4$), has showed that the achieved performance is worse than that exposed in Section~\ref{sec:perf-results}.

Given the constraints on polyhedral compilation and linear-algebra-specialized libraries, we address assembly optimization by studying a set of domain-specific code transformations, applicable to a wide class of problems. We have developed a domain-specific optimizing compiler and we have integrated it with Firedrake, a system for solving PDEs through the Finite Element Method~\cite{firedrake} based on the OP2 abstraction. This allows us to evaluate our code transformations in a range of real-world problems, varying parameters that impact both solution accuracy and kernel cost, namely the polynomial order of the method (from $p=1$ to $p=4$) and the mesh type (2D, 3D, 3D hybrid structured-unstructured).

Performance results show that Firedrake-generated code for non-trivial assembly kernels, if not adequately optimized, is sub-optimal. A cost-model-driven sequence of source-to-source code transformations, aimed at improving register allocation and register data locality, can result in performance improvements up to 33$\%$ over ``softly-optimized'' code (i.e. where only basic transformations are performed, such as auto-vectorizable loop-invariant code motion, padding and data alignment), and up to 77$\%$ over the original kernels. The contribution of this paper is therefore twofold:
\begin{itemize}
\item An optimisation strategy for a class of kernels widely-used in scientific applications, namely Local Assembly in the context of the Finite Element Method. Our approach exploits domain knowledge to go beyond the limits of both vendor (e.g. $icc$) and research (e.g. those based on the polyhedral model) compilers.
\item Automation of such code optimizations in Firedrake. This enables performing an in-depth performance evaluation of the proposed optimization strategy in real-world simulations.
\end{itemize}

@@

\section{Background}
%Here we discuss about the computational characteristics of computational science kernels. Briefly cite op2 kernels. Emphasis on Finite Element Assembly. Generalization and formalization of a Finite Element Assembly kernel using quadrature representation.
\begin{algorithm}[t]
\small
\caption{Local assembly code generated by Firedrake for a Helmholtz problem (2D mesh, Lagrange $p=1$ elements).}
\label{code:helmholtz}
\KwSty{void} helmholtz(double A[3][3], double **coords) $\lbrace$\\
~~// K, det = Compute Jacobian (coords) \\
~~\\
~~\KwSty{static const double} W3[3] = $\lbrace$...$\rbrace$\\
~~\KwSty{static const double} X$\_$D10[3][3] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\KwSty{static const double} X$\_$D01[3][3] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$3; i++) \\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$3; k++) \\
~~~~~~~~A[j][k] += ((Y[i][k]*Y[i][j]+\\
~~~~~~~~~~~+((K[1]*X$\_$D10[i][k]+K[3]*X$\_$D01[i][k])*\\
~~~~~~~~~~~*(K[1]*X$\_$D10[i][j]+K[3]*X$\_$D01[i][j]))+\\
~~~~~~~~~~~+((K[0]*X$\_$D10[i][k]+K[2]*X$\_$D01[i][k])*\\
~~~~~~~~~~~*(K[0]*X$\_$D10[i][j]+K[2]*X$\_$D01[i][j])))*\\
~~~~~~~~~~~*det*W3[i]);\\
$\rbrace$
\end{algorithm}

@@

Local assembly consists of evaluating so called element stiffness matrix and vector; in this work, we focus on computation of matrices, which is the costly part of the process. A stiffness matrix can be intuitively thought as an approximated representation of the PDE solution in a specific cell of the discretized domain. Numerical integration algorithms are widely-used in scientific codes to implement assembly~\cite{quadrature1, quadrature2}.

Given a mathematical description of the input problem, expressed through the domain-specific Unified Form Language~\cite{ufl}, Firedrake generates C-code kernels implementing assembly using a numerical integration algorithm. At the core of Firedrake~\cite{firedrake} lies PyOP2~\cite{pyop2ics}, which triggers compilation of such kernels using an available vendor compiler, and eventually manages parallel execution over the mesh. In our work, we have enhanced this execution model by adding an optimization stage prior to the generation of C code. The code transformations described next are also generalizable to non-Firedrake assembly kernels, provided that numerical integration is used. Notable examples are~\cite{uflacs, fluidity, mcfc}, in which the kernel's loop nest and memory access pattern are reducible to that generated by Firedrake.

The complexity of Firedrake-generated kernels depends on the mathematical problem being solved. In simpler cases, the loop nest is perfect, it has short trip counts (in the range 3-15), and the computation reduces to a summation of a few, small outer-product vector-operations. An example is provided in Figure~\ref{code:helmholtz}, which shows an assembly kernel for a Helmholtz problem, using Lagrange basis functions on 2D elements with polynomial order $p=1$. In other scenarios, for instance when solving a non-linear problem like Burgers (see Figure~\ref{code:burgers}), the number of arrays involved in the computation of $A$ can be much larger: in this case, 14 unique arrays are accessed, and the same array can be referenced multiple times within the expression. Also, constants evaluated in outer loops (called $F$ in the code), acting as scaling factors of arrays, may be required; trip counts can be larger (proportionally to the order of the method); arrays may be block-sparse. Despite such a big space of possible assembly implementations, it is still possible to identify a domain-specific structure common to all kernels, which can be turned into effective code transformations and simd-vectorization.

@@

@@

\begin{algorithm}[t]
\small
\caption{Local assembly code generated by Firedrake for a Helmholtz problem (2D mesh, Lagrange $p=1$ elements). The padding, data alignment and $licm$ optimizations are applied. Data alignment and padding relate to an AVX machine. In this specific case, sub-expressions invariant to $j$ are identical to those invariant to $k$; in general, this could not be the case.}
\label{code:helmholtz-licm}
\KwSty{void} helmholtz(double A[3][4], double **coords) $\lbrace$\\
~~// K, det = Compute Jacobian (coords) \\
~~\\
~~\KwSty{static const double} W3[3] = $\lbrace$...$\rbrace$\\
~~\KwSty{static const double} X$\_$D10[3][4] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\KwSty{static const double} X$\_$D01[3][4] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$3; i++) $\lbrace$ \\
~~~~double LI$\_$0[4];\\
~~~~double LI$\_$1[4];\\
~~~~\KwSty{for} (\KwSty{int} r = 0; r$<$4; r++) $\lbrace$ \\
~~~~~~LI$\_$0[r] = ((K[1]*X$\_$D10[ip][r])+(K[3]*X$\_$D01[ip][r]));\\
~~~~~~LI$\_$1[r] = ((K[0]*X$\_$D10[ip][r])+(K[2]*X$\_$D01[ip][r]));\\
~~~~$\rbrace$\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
~~~~~~~~A[j][k] += (Y[i][k]*Y[i][j]+LI$\_$0[k]*LI$\_$0[j]+\\
~~~~~~~~~~~~~~~~~~~~+LI$\_$1[k]*LI$\_$1[j])*det*W3[i]);\\
~~$\rbrace$\\
$\rbrace$
\end{algorithm}

\begin{algorithm}[t]
\small
\caption{Local assembly code generated by Firedrake for a Helmholtz problem (2D mesh, Lagrange $p=1$ elements). The padding, data alignment, $licm$ and $split$ optimizations are applied. Data alignment and padding relate to an AVX machine. In this specific case, the split factor is 2.}
\label{code:helmholtz-split}
\KwSty{void} helmholtz(double A[3][4], double **coords) $\lbrace$\\
~~// K, det = Compute Jacobian (coords) \\
~~// Declaration of basis function matrices \\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$3; i++) $\lbrace$ \\
~~~~double LI$\_$0[4];\\
~~~~double LI$\_$1[4];\\
~~~~\KwSty{for} (\KwSty{int} r = 0; r$<$4; r++) $\lbrace$ \\
~~~~~~LI$\_$0[r] = ((K[1]*X$\_$D10[ip][r])+(K[3]*X$\_$D01[ip][r]));\\
~~~~~~LI$\_$1[r] = ((K[0]*X$\_$D10[ip][r])+(K[2]*X$\_$D01[ip][r]));\\
~~~~$\rbrace$\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
~~~~~~~~A[j][k] += (Y[i][k]*Y[i][j]+LI$\_$0[k]*LI$\_$0[j])*det*W3[i];\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
~~~~~~~~A[j][k] += LI$\_$1[k]*LI$\_$1[j]*det*W3[i];\\
~~$\rbrace$\\
$\rbrace$
\end{algorithm}

@@

\begin{algorithm}[t]
\small
\caption{Local assembly code generated by Firedrake for a Helmholtz problem (2D mesh, Lagrange $p=2$ elements). The padding, data alignment, $licm$ and $op$-$vect$ optimizations are applied. Data alignment and padding relate to an AVX machine. In this specific case, the unroll-and-jam factor is 1.}
\label{code:helmholtz-opvect}
\KwSty{void} helmholtz(double A[8][8], double **coords) $\lbrace$\\
~~// K, det = Compute Jacobian (coords) \\
~~// Declaration of basis function matrices \\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$6; i++) $\lbrace$ \\
~~~~// Do loop-invariant code motion \\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$4; j+=4) \\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$8; k+=4) $\lbrace$\\
~~~~~~~~// Call $Load$ and $set$ intrinsics \\
~~~~~~~~// Compute A[1,1],A[2,2],A[3,3],A[4,4] \\
~~~~~~~~// One $permute\_pd$ intrinsics per $k$-loop $load$\\
~~~~~~~~// Compute A[1,2],A[2,1],A[3,4],A[4,3] \\
~~~~~~~~// One $permute2f128\_pd$ intrinsics per $k$-loop $load$\\
~~~~~~~~// ...\\
~~~~~~$\rbrace$\\
~~~~// Do Remainder loop (from $j=4$ to $j=6$)\\
~~$\rbrace$\\
~~// Restore the storage layout:\\
~~\KwSty{for} (\KwSty{int} j = 0; j$<$4; j+=4) $\lbrace$\\
~~~~$\_\_$m256d r0, r1, r2, r3, r4, r5, r6, r7;\\
~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$8; k+=4) $\lbrace$\\
~~~~~~r0 = $\_$mm256$\_$load$\_$pd ($\&$A[j+0][k]);\\
~~~~~~// Load A[j+1][k], A[j+2][k], A[j+3][k]\\
~~~~~~r4 = $\_$mm256$\_$unpackhi$\_$pd (r1, r0);\\
~~~~~~r5 = $\_$mm256$\_$unpacklo$\_$pd (r0, r1);\\
~~~~~~r6 = $\_$mm256$\_$unpackhi$\_$pd (r2, r3);\\
~~~~~~r7 = $\_$mm256$\_$unpacklo$\_$pd (r3, r2);\\
~~~~~~r0 = $\_$mm256$\_$permute2f128$\_$pd (r5, r7, 32);\\
~~~~~~r1 = $\_$mm256$\_$permute2f128$\_$pd (r4, r6, 32);\\
~~~~~~r2 = $\_$mm256$\_$permute2f128$\_$pd (r7, r5, 49);\\
~~~~~~r3 = $\_$mm256$\_$permute2f128$\_$pd (r6, r4, 49);\\
~~~~~~$\_$mm256$\_$store$\_$pd ($\&$A[j+0][k], r0);\\
~~~~~~// Store A[j+1][k], A[j+2][k], A[j+3][k]\\
~~~~$\rbrace$\\
~~$\rbrace$\\
$\rbrace$
\end{algorithm}

@@

\begin{table}[h]
\begin{tabulary}{1.0\columnwidth}{|C|C|}
\hline
Name (Abbreviation) & Parameter \\\hline\hline
Auto-vectorizable Loop-invariant code motion ($licm$) &   \\ \hline
Padding &  \\ \hline
Data Alignment & \\ \hline
Loop interchange      & loops  \\ \hline
Loop unrolling  & unroll factor \\ \hline
Register tiling & tile size \\ \hline
Outer-product vectorization ($op$-$vect$) & tile size \\ \hline
Assembly splitting ($split$) & split point, split factor \\ \hline
\end{tabulary}
\caption{Overview of code transformations for Firedrake-generated assembly kernels.}
\label{table:code-transformations}
\end{table}

@@

Firedrake provides users with the Unified Form Language to write problems in a notation resembling mathematical equations. This high-level specification is translated by the domain-specific Fenics Form Compiler~\cite{ffc} into an Abstract Syntax Tree representation of a Finite Element assembly kernel. ASTs are then passed to PyOP2, where parallel execution over the unstructured mesh is managed. Our compiler, capable of applying the transformations described in Section~\ref{sec:code-transf}, is integrated with PyOP2: it can manipulate ASTs introducing optimizations, and then it generates C code, which is eventually just-in-time compiled on the underlying architecture.
